---
title: "Text Gen in the Classroom"
subtitle: "Understanding student essays in the CIC corpus"
author: "Brighton Pauli"
---

# Imports and data

```{r, message=FALSE}
library("tidyverse")    # data wrangling and visualization
library("magrittr")     # %>% pipe
library("ggplot2")      # fancy data visualization
library("lme4")         # linear mixed effects models
library("forcats")      # reordering for bar plot

data_dir <- "D:/data/claim_identification_corpus"
```

```{r}
df <- "EssaysWithPPL.csv" %>%
    sprintf("%s/%s", data_dir, .) %>%
    read.csv
df <- filter(df, LinePPL < 1000)
glimpse(df)
```

```{r}
valid_types <- c("Concluding_Summary", "Counterclaim", "Data", "Final_Claim",
                 "Nonannotated", "Primary_Claim", "Rebuttal")
df_annot <- "AnnotsWithPPL.csv" %>%
    sprintf("%s/%s", data_dir, .) %>%
    read.csv %>%
    filter(ppl < 600) %>%
    filter(adjudicated_discourse_type %in% valid_types)
glimpse(df_annot)
```

# Entire Essays

## Visualizing perplexity on its own

```{r}
hist(df$LinePPL)
```

## Directly comparing perplexity and scores

Holistic score based solely on average line perplexity.

```{r}
score_ppl <- lm(Holistic.Score ~ LinePPL, data = df)
summary(score_ppl)
```

```{r}
ggplot(df, aes(LinePPL, Holistic.Score)) +
geom_point()
```

Average line perplexity based on the total length. Colored by holistic score.

```{r}
length_ppl <- lm(LinePPL ~ Number.of.words, data = df)
summary(length_ppl)
```

```{r}
ggplot(df, aes(Number.of.words, LinePPL, color = Holistic.Score)) +
geom_point() +
scale_color_gradient2(low = "#1a1c37", mid = "#818285", high = "#fac864",
    midpoint = 3) +
geom_smooth(method = "lm", color = "#6496c8")
```

## Categorical scores and perplexities

```{r}
df <- df %>%
    mutate(ppl_label = case_when(                LinePPL < 60 ~ "AI",
                                 60 <= LinePPL & LinePPL < 80 ~ "Unsure",
                                 80 <= LinePPL            ~ "Human"))
glimpse(df)
```

```{r}
df_summ <- df %>%
    group_by(Holistic.Score, ppl_label) %>%
    summarize(n = n()) %>%
    mutate(freq = n / sum(n))
df_summ$ppl_label <- factor(df_summ$ppl_label,
                            c("AI", "Unsure", "Human"))
glimpse(df_summ)
```

```{r}
df_summ %>%
ggplot(aes(x = Holistic.Score, fill = ppl_label, group = ppl_label)) +
geom_bar(aes(y = freq), stat = "identity") +
scale_fill_manual(values = c("#e85d75", "#818285", "#6469c8")) +
labs(
    x = "Holistic Score",
    y = "Frequency",
    fill = "GPTZero Label"
) +
ggtitle("Probability of False Positive by Holistic Score")
```

```{r}
df_summ %>%
ggplot(aes(x = ppl_label, fill = Holistic.Score, group = Holistic.Score)) +
theme_light() +
geom_bar(aes(y = freq), stat = "identity")
```

# Individual Annotations

## Visualizing Perplexities

```{r}
hist(df_annot$ppl, breaks = 30, xlim = c(0, 150), xlab = "Perplexity",
     main = "Histogram of Perplexities")
```

## Linear Regression

```{r}
lm(ppl ~ score, data = df_annot) %>%
    summary
```

```{r}
ggplot(df_annot, aes(x = score, y = ppl)) +
geom_point() +
geom_smooth(method = "lm") +
xlab("Score") +
ylab("Perplexity")
```

## Linear Mixed Effects Regression

```{r}
score_mixed <- lmer(
    ppl ~ score + (1 | adjudicated_discourse_type),
    data = df_annot)
score_mixed
```

## Perplexity per Discourse Type

Discourse elements should differ in perplexity. High-entropy elements like
"Data" vary more than low-entropy elements like "Primary_Claim"s, since there
should only be two types of primary claims ("I agree" or "I disagree"), but 
there are countless potential examples.

```{r}
ggplot(df_annot,
    aes(y = ppl, x = adjudicated_discourse_type,
    fill = adjudicated_discourse_type)) +
geom_boxplot() +
theme(legend.position = "none") +
xlab("Discourse Type") +
ylab("Perplexity")
```

### With Violins

```{r}
df_annot %>%
ggplot(aes(
    y = ppl,
    x = adjudicated_discourse_type,
    fill = adjudicated_discourse_type)) +
geom_boxplot() +
geom_violin(alpha = 0.2, ) +
theme(legend.position = "none") +
xlab("Discourse Type") +
ylab("Perplexity")
```

### Histogram of just Data

```{r}
df_annot %>%
    filter(adjudicated_discourse_type == "Data") %>%
    ggplot(aes(x = ppl)) + geom_histogram()
```
